Mathematical Approaches for Simulating Epidemic Progression: Addressing Limitations of the Linear Chain Trick in ODE Models

### 2024-03-27 thesis/draft.tex
temp storage

To illustrate this point, consider that the Gamma distribution can be parameterized in terms of its mean $(D)$ and shape parameter (quadratic coefficient of variation, $\kappa$), which determine the overall shape and position of the distribution. As mentioned above, in subdividing a single stage into n substages, each exponentially distributed substage must have a rate of $n\gamma$ to maintain the overall rate of that stage. While the mean of the Erlang distribution in the ODE system can be chosen freely, $\kappa$ is constrained to discrete values, shown as follows:
\begin{align*}
    Mean: \quad &\sum_{i=1}^{n_f} \frac{1}{n\gamma} = \frac{1}{\gamma}\\
    Variance: \quad &\sum_{i=1}^{n_f} \frac{1}{n^2\gamma^2} = \frac{1}{n\gamma^2} \\
    \kappa: \quad &\frac{\sigma^2}{M^2} = \frac{1}{n} \quad n \in \mathbb{N}
\end{align*}
Our approach, using a geometric series of rates, addresses this issue by allowing $\kappa$ to take any value within the range $\left[\ \frac{1}{n}, 1 \right)$. A detailed explanation and the derivation process are provided in section $4.2$. Another advantage of this method is that it enables a fixed model structure where only the parameters change, eliminating the need to determine optimal substage numbers through the construction of multiple models, a process currently required in the ODE model that employs the LCT.

@@

The dynamic of the population within the three compartment can be expressed as a system of nonlinear ordinary differential equations:
\begin{align}
    \frac{dS}{dt} &= \mu N - \frac{\beta SI}{N} - \mu S \\
    \frac{dI}{dt} &= \frac{\beta SI}{N} - \gamma I - \mu I \\
    \frac{dR}{dt} &= \gamma I - \mu R
\end{align}
where $S$, $I$, and $R$ represent the susceptible, recovered, and infectious population. $N$ is the size of the total population, and in this case $S + I + R = N$. The $\mu$, $\beta$ and $\gamma$ are the rates of natural birth (death), transmission and recovery, respectively. $\beta$ is the rate of new infections per unit time, so $\beta SI$ is the number of new infections that occur per unit time (the incidence rate). Here, the mean recovery time for an infected individual is $\frac{1}{\gamma}$. By assuming that the entire population was infected at the onset and considering a zero natural birth (death) rate, we can focus solely on the dynamics of the recovered population, as indicated by formula (1).

@@

\section*{Results}

>>>>

To illustrate this point, consider that the Gamma distribution can be parameterized in terms of its mean $(D)$ and shape parameter (quadratic coefficient of variation, $\kappa$), which determine the overall shape and position of the distribution. As mentioned above, in subdividing a single stage into n substages, each exponentially distributed substage must have a rate of $n\gamma$ to maintain the overall rate of that stage. While the mean of the Erlang distribution in the ODE system can be chosen freely, $\kappa$ is constrained to discrete values, shown as follows:
\begin{align*}
    Mean: \quad &\sum_{i=1}^{n_f} \frac{1}{n\gamma} = \frac{1}{\gamma}\\
    Variance: \quad &\sum_{i=1}^{n_f} \frac{1}{n^2\gamma^2} = \frac{1}{n\gamma^2} \\
    \kappa: \quad &\frac{\sigma^2}{M^2} = \frac{1}{n} \quad n \in \mathbb{N}
\end{align*}
Our approach, using a geometric series of rates, addresses this issue by allowing $\kappa$ to take any value within the range $\left[ \frac{1}{n}, 1 \right)$. A detailed explanation and the derivation process are provided in section $4.2$. Another advantage of this method is that it enables a fixed model structure where only the parameters change, eliminating the need to determine optimal substage numbers through the construction of multiple models, a process currently required in the ODE model that employs the LCT.

@@

The dynamic of the population within the three compartment can be expressed as a system of nonlinear ordinary differential equations:
\begin{align}
    \frac{dS}{dt} &= \mu N - \frac{\beta SI}{N} - \mu S \\
    \frac{dI}{dt} &= \frac{\beta SI}{N} - \gamma I - \mu I \\
    \frac{dR}{dt} &= \gamma I - \mu R
\end{align}
where $S$, $I$, and $R$ represent the susceptible, recovered, and infectious population. $N$ is the size of the total population, and in this case $S + I + R = N$. The $\mu$, $\beta$ and $\gamma$ are the rates of natural birth (death), transmission and recovery, respectively. $\beta$ is the rate of new infections per unit time, so $\beta SI$ is the number of new infections that occur per unit time (the incidence rate). Together, formulas $(3,4,5)$ will be referred to as the transmission model. Here, the mean recovery time for an infected individual is $\frac{1}{\gamma}$. By assuming that the entire population was infected at the onset and considering a zero natural birth (death) rate, we can focus solely on the dynamics of the recovered population, as indicated by formula (1).

@@

After confirming that the GCT model, which employs a fixed structure and varying parameters $a$ and $r$, achieves comparable efficacy to the LCT model, we proceeded to empirically test the model's enhanced performance. Specifically, we focused on its improved flexibility in the kappa parameter through data fitting. Initially, we utilized simulated data as a replacement for actual data. This incidence data was generated through transmission model simulations, with deviations (noise) introduced via a mean reporting ratio and a negative binomial dispersion parameter. We performed model fitting using the \textbf{bbmle} package in R, estimating the model parameters through Maximum Likelihood Estimation (MLE). Our goal was to accurately retrieve the parameter values used to generate the incidence data, assessing the fitting quality through the mean squared error (MSE).

We began by applying the same model type (LCT or GCT) for both data generation and model fitting. Data were generated by simulating the LCT model with two substages $(n=2)$ and then fitted using the LCT model, evaluating both identical and modified structures. This methodology was similarly applied to the GCT model. To further test the GCT model, we increased the fitting challenge by generating incidence data using the LCT model and then fitting it with the GCT model, varying the number of substages. Subsequently, we modified both the fitting model's structure and the input data to observe the impact on fitting outcomes. Finally, we extracted the parameters estimated through fitting and compared them with the actual values (those used for simulation).

In the next phase of our study, we will approach the fitting of interval data from two perspectives: the rounding paradigm, which interprets 2 days as meaning any duration between 1.5 and 2.5 days, and the start/end paradigm, which considers potential starting and ending times. Initially, we will generate interval data using various distributions, such as gamma and lognormal, and subsequently apply both the LCT and GCT models for fitting. Although the GCT model can replicate certain properties (such as $D$ and $\kappa$) observed with the gamma distribution, its inherent distribution does not match the gamma distribution precisely. Thus, it's necessary for us to develop a new likelihood function for fitting purposes. Furthermore, we intend to incorporate real-world data, including incidence and interval data, to evaluate the GCT model's performance comprehensively. By comparing the fitting results from both the LCT and GCT models, we aim to quantitatively determine the effectiveness of our proposed method in enhancing model accuracy and its reflection of actual epidemiological scenarios.

\section{Results}
\subsection{Independent and Flexible Modification of Mean and Kappa in Stage Duration Distribution Using the GCT Model}
As discussed in previous sections, the duration distribution of the stage can be parameterized by its mean (D) and shape parameter $(\kappa^2 = CV^2)$. In the GCT model, the calculation of mean and variance is achieved by summing the respective properties of each substage, shown as follows:
\begin{align}
    Mean: \quad  &\sum_{i=1}^{n_f} \frac{1}{\gamma_i} = \frac{\frac{1}{a} (\frac{1}{r^n} - 1)}{\frac{1}{r} -1} \\
    Variance: \quad &\sum_{i=1}^{n_f} \left(\frac{1}{\gamma_i} \right)^2 = \frac{\frac{1}{a^2} (\frac{1}{r^{2n}} - 1)}{\frac{1}{r^2} -1}
\end{align}
The shape parameter $\kappa$, determined by $\frac{\sigma^2}{D}$, can thus be obtained:
\begin{align}
    \kappa = \frac{(1+r^{n}) (1-r)}{(1+r) (1-r^{n})}
\end{align}
This simplification revealed that the value of $\kappa$ is solely determined by the parameter $r$, thus enabling the alignment of kappa without affecting the mean, which can be adjusted through alterations to parameter $a$. Similarly, the mean can be scaled to a desired value without impacting $\kappa$. This grants the model independence in modifying properties.

Importantly, within the GCT model, even though the number of substages can be modified $(n)$, the value of $\kappa$ becomes independent of this number. It should also be noted that, according to formula (20), the range of $\kappa$ can encompass any real number within the range $\left[ \frac{1}{n}, 1 \right)$ for $r > 0$, thereby significantly enhancing the model's flexibility. In contrast, when applying a similar calculation for $\kappa$ in the LCT model, we find that $\kappa = \frac{1}{n}$ and the shape parameter is therefore constrained to a series of discrete values, as follows:
\begin{align*}
    \kappa \in \left\{ 1, \frac{1}{2}, \frac{1}{3}, \dots \right\}
\end{align*}
The transition from discrete to continuous values in kappa significantly enhances the model's adaptability to the shape of the stage duration distribution, enabling a more precise fit to the observed data and allowing for a nuanced interpretation of temporal patterns.

\subsection{Matching GCT and LCT Model Performance through Parameter Adjustment within a Uniform Structure}

Using formulas (18) and (20), we developed code to reverse-engineer the process: given specific values of D and kappa, we can identify the corresponding $a$ and $r$ that endow the model with such characteristics. These $a$ and $r$ values were subsequently integrated into the GCT model. Our initial goal was to determine if the stage duration distribution could match between the two models, assuming identical $D$ and $\kappa$ values. The number of substages in the LCT model was set to 2, 4, 6, and 8, while the substage number for the GCT model was fixed at 12. We selected a mean of 10 days. Applying the mean and kappa values $\left( \frac{1}{2}, \frac{1}{4}, \frac{1}{6}, \frac{1}{8} \right)$ to the system, we plotted the stage duration distributions generated from the LCT cohort model alongside those from the GCT with the corresponding a and r values. As depicted in Figure x.x, the stage duration distributions exhibit slightly different shapes, despite having identical mean and shape parameters. To further examine the match of properties, D and kappa were numerically calculated from the output distributions, revealing identical values across both models. This indicates that while the distributions' shapes may not completely match, they can possess the same properties.

By reintegrating the infectious stage duration distribution into the whole dynamic system, we present the dynamics of newly infected cases (incidence) as simulated by the transmission model. Figure x.x illustrates the incidence time series for the LCT model with 4 and 8 substages, represented by yellow and blue lines, respectively. The corresponding kappa values of the stage duration distribution for them are $\frac{1}{4}$ and $\frac{1}{8}$. Employing the GCT model with a fixed 12 substages and utilizing $a$ and $r$ values derived from the reverse-engineering system for the targeted kappa values, we successfully approximate the incidence patterns observed in the LCT model across different substage configurations, as shown by the black dashed line. This demonstrates the GCT model's capability to emulate the LCT model's performance by adjusting parameters within a constant structural framework. Compared to modifying the model structure to achieve desired kappa values, the GCT model simplifies parameter estimation and model fitting processes.

\subsection{Substage Configuration Minimally Impacts Model Fitting}
\textbf{Only temporarily. I'm going to change all of this...}

After establishing the theoretical foundation, we proceeded with model fitting to assess the GCT model's performance. The incidence data were obtained from the model simulation, and the noises were added using a mean reporting ratio of 0.9 and a negative binomial dispersion parameter of 1000. Initially, we employed the same model type for both data generation and fitting. In Figure x.x.A, the data, generated by the LCT model with 2 substages, were fitted using an LCT model of the identical structure, as the plot title indicates. The quality of the fitting was evaluated based on the mean squared error (MSE). With an MSE of 1.2, the fitting can be considered satisfactory. Figure x.x.B showcases the parameters estimated post-fitting. Conversely, employing the same incidence data but altering the fitting model's structure from an LCT with 2 substages to an LCT with 6 substages, as depicted in Figure x.x.C, resulted in less favorable outcomes, with the MSE escalating to 25. The estimated parameters, displayed in Figure x.x.D, showed significant discrepancies.

On the other hand, employing the GCT model with 12 substages for both generating and fitting incidence data resulted in a satisfactory fit, as indicated by an MSE of 0.01 (Figure x.x.A), and parameters post-fitting were close to the exact values (Figure x.x.B). However, altering the fitting model's structure from 12 substages to 6 substages, the model still exhibited a close match with the original in terms of both shape (Figure x.x.C) and parameters (Figure x.x.D), albeit with a slightly increased MSE (Figure x.x.C). This indicates that the GCT model's fitting performance is minimally impacted by changes in the model structure.

\subsection{Adaptability of the GCT Model to Data Input and Structural Changes in Fitting}
Proceeding to a more challenging test, we utilized a different model type for data generation and model fitting. As demonstrated in Figure x.x, the data generated by the LCT model with 2 substages were fitted using the GCT model with 12 substages. Despite the significant difference in model structure, the fitting results were reasonably favorable, with an MSE of 2.4. Altering the fitting model from a GCT with 2 substages to a GCT with 6 substages had negligible impact on the fitting outcomes (Figure x.x). Similarly, changing the input data from an LCT with 2 substages to an LCT with 7 substages resulted in slightly improved fitting, though the difference was not substantial (Figure x.x). These findings suggest that regardless of changes to the fitting model structure or the data used, the GCT model is capable of adapting to these variations, providing satisfactory fitting results.

\subsection{Parameter Estimation}
In this section, we discuss the parameter estimation results obtained through model fitting. The incidence data were generated using the GCT model with 6 substages, setting the $D$ to 10 and $\kappa$ to $\frac{2}{9}$. The model was fitted using the GCT framework with 2 to 10 substages. As depicted in Figure x.x, the precise values for each parameter are marked by a red line, with post-fitting parameter values displayed alongside their 95\% confidence intervals. It can be observed that the parameter estimation is not perfect, with kappa and beta consistently overestimated, and D underestimated. However, these discrepancies in parameter estimation counterbalance each other, resulting in a fitting outcome characterized by a low mean squared error. The absence of confidence intervals for models with 2 and 4 substages is attributed to structural inadequacy. As previously mentioned, the minimum possible value for kappa in the model is $\frac{1}{n}$. The model encounters limitations in aligning with a kappa value of 2/9 when using 2 and 4 substages. It is important to acknowledge the imprecision of parameter estimation, which will be a focus of our subsequent efforts.
